---
title: "Danish Monarchs Data R Markdown"
author: "Julie"
date: "5/8/26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#I first need to load the tidyverse packages again
library(tidyverse)
library(ggplot2)
```


The task here is to load your monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the monarchs

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `monarchs.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator:_______  

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv("data/")`
-  `read_csv("data/")`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=FALSE}
# FILL IN THE CODE BELOW and review the outputs
monarchs1 <- read.csv("data_monarchs/danish_monarchs_data.csv")

monarchs2 <- read_csv("data_monarchs/danish_monarchs_data.csv")

monarchs3 <- read.csv2("data_monarchs/danish_monarchs_data.csv")

monarchs4 <- read_csv2("data_monarchs/danish_monarchs_data.csv")
```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `monarchs` object
#Answer: I used read_csv2 which is semicolon-delimited files.

2. What is the result of running `class()` on the `monarchs` object created with a tidyverse function.
#Answer: It tells me that the monarchs object is: A tibble (tbl_df, tbl), the tidyverse's enhanced data frame, till a data.frame at its core, as a spec_tbl_df class (spec = specification - tracks column types)

3. How many columns does the object have when created with these different functions?
#Answer: The object now has 5 columns: Name, Birth_year, Death_year, Reign_start, Reign_end.

4. Show the dataset so that we can see how R interprets each column
#Answer: To view my data I used the functions class, glimpse and head as seen below.

```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

monarchs <- read_csv2("data_monarchs/danish_monarchs_data.csv")

class(monarchs)

glimpse(monarchs)

head(monarchs)


```


## Calculate the duration of reign for all the monarchs in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the monarchs dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r eval = TRUE}
# YOUR CODE
monarchs <- monarchs %>%
  mutate(duration = Reign_end - Reign_start)
```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `monarchs` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r eval=TRUE}
# YOUR CODE
monarchs <- monarchs %>%
  mutate(duration = as.numeric(Reign_end) - as.numeric(Reign_start))

mean(monarchs$duration, na.rm = TRUE)

#Answer to the question is 20.83019 years as average duration of reign
```

## How many and which monarchs enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `monarchs` dataset. Display the result and also  count the resulting rows with `count()`

```{r eval=TRUE}
# YOUR CODE
monarchs <- monarchs %>%
  mutate(duration = as.numeric(Reign_end) - as.numeric(Reign_start))

mean(monarchs$duration, na.rm = TRUE)

monarchs %>%
  filter(duration > mean(monarchs$duration, na.rm = TRUE)) %>%
  count()

# Answer is 24
```


## How many days did the three longest-ruling monarchs rule? 

* Sort monarchs by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r eval=TRUE}
# YOUR CODE
#finding the three longest-ruling monarchs and converting their reign duration from years to days.

monarchs %>%
  arrange(desc(duration)) %>%
  slice(1:3) %>%
  mutate(days = duration * 366) %>%
  select(Name, duration, days)

#Answer is Erik 7. af Pommern, Christian 4. og Margrethe 2.
```


## Challenge: Plot the monarchs' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r eval=TRUE}
# YOUR CODE

monarchs %>%
  ggplot(aes(x = Reign_start, y = duration)) +
  geom_point() +
  geom_line() +
  labs(title = "Duration of Reign Through Time",
       x = "Start of Reign (Year)",
       y = "Duration (Years)") +
  theme_minimal()
```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!